---
title: "Analyze DE Gene and SJ by Cell type"
output: html_document
date: "2024-01-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The purpose of this script is to analyze differential analysis results MARVEL. It is dependent on running all seurat scripts, and marvel scripts 01 and 02. Please use docker image setbp1_alternative_splicing:1.0.5.

## Load Packages

```{r load packages}
suppressPackageStartupMessages({
  # Load Lasseigne Lab standard packages
  library(here)
  library(styler)
  library(lintr)
  library(tidyverse)
  library(patchwork)

  # Load MARVEL package
  library(MARVEL)

  # Load adjunct MARVEL packages
  library(ggnewscale)
  library(ggrepel)
  library(reshape2)
  library(plyr)
  library(stringr)
  library(textclean)
  library(AnnotationDbi)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(org.Mm.eg.db)
  library(gtools)
  library(GenomicRanges)
  library(IRanges)
  library(S4Vectors)
  library(wiggleplotr)
  library(Matrix)
  library(data.table)
  library(gridExtra)
})

# set seed
set.seed(123)

# source functions
source(here::here("src", "marvel", "functions.R"))
```


## Gene-splicing dynamics visalization - excitatory neurons

```{r what are the significant genes for this analysis}
# Pull significant genes
excitatory_sig_table <-
  excitatory_marvel[["SJ.Gene.Cor"]][["Data"]]

# Pull significant genes
excitatory_sig_genes <-
  excitatory_marvel[["SJ.Gene.Cor"]][["Data"]]$gene_short_name

# Pull SJ locations
excitatory_sig_coords <-
  excitatory_marvel[["SJ.Gene.Cor"]][["Data"]]$coord.intron
```


```{r run umap plotting function on all excitatory genes}
excitatory_marvel <- plot_marvel_umap(
  marvel_object = excitatory_marvel,
  prefix = "excitatory",
  gene = "Kif21a",
  sj_loc = "chr15:90881929:90882035"
)

excitatory_marvel <- plot_marvel_umap(
  marvel_object = excitatory_marvel,
  prefix = "excitatory",
  gene = "Pnisr",
  sj_loc = "chr4:21859308:21860819"
)

excitatory_marvel <- plot_marvel_umap(
  marvel_object = excitatory_marvel,
  prefix = "excitatory",
  gene = "Igf1r",
  sj_loc = "chr7:67602701:67653557"
)

excitatory_marvel <- plot_marvel_umap(
  marvel_object = excitatory_marvel,
  prefix = "excitatory",
  gene = "Luc7l3",
  sj_loc = "chr11:94200518:94212515"
)

excitatory_marvel <- plot_marvel_umap(
  marvel_object = excitatory_marvel,
  prefix = "excitatory",
  gene = "Phf14",
  sj_loc = "chr6:11926837:11933251"
)

excitatory_marvel <- plot_marvel_umap(
  marvel_object = excitatory_marvel,
  prefix = "excitatory",
  gene = "Calm1",
  sj_loc = "chr12:100172493:100172658"
)

excitatory_marvel <- plot_marvel_umap(
  marvel_object = excitatory_marvel,
  prefix = "excitatory",
  gene = "Son",
  sj_loc = "chr16:91448685:91451498"
)

excitatory_marvel <- plot_marvel_umap(
  marvel_object = excitatory_marvel,
  prefix = "excitatory",
  gene = "Afdn",
  sj_loc = "chr17:14024397:14027538"
)

excitatory_marvel <- plot_marvel_umap(
  marvel_object = excitatory_marvel,
  prefix = "excitatory",
  gene = "Srrm2",
  sj_loc = "chr17:24027148:24027327"
)

```

## Gene Ontology Analysis

Unfortunately, at least 10 genes are required for GO analysis, so I am skipping this.

## Candidate Gene Analysis

It looks like I can make as many cell groups and label as I like. This may help us for comparing expression of a gene or splice junction across all cell types for both mutant and control.

### Plot SJ position

```{r plot SJ position}
#install.packages(here::here("bin", "custom_packages", "wiggleplotr_1.18.0.tar.gz"), repos = NULL, type = "source")

excitatory_marvel <- adhocGene_PlotSJPosition_EJ(
  MarvelObject = excitatory_marvel,
  coord.intron = "chr15:90881929:90882035",
  rescale_introns = FALSE,
  show.protein.coding.only = TRUE,
  anno.label.size = 1.5
)

excitatory_marvel$adhocGene$SJPosition$Plot
```

## Gene-splicing dynamics visalization - inhibitory neurons

```{r what are the significant inhibitory genes}
# Pull significant genes
inhibitory_sig_table <-
  inhibitory_marvel[["SJ.Gene.Cor"]][["Data"]]

# Pull significant genes
inhibitory_sig_genes <-
  inhibitory_marvel[["SJ.Gene.Cor"]][["Data"]]$gene_short_name

# Pull SJ locations
inhibitory_sig_coords <-
  inhibitory_marvel[["SJ.Gene.Cor"]][["Data"]]$coord.intron
```


```{r what are the significant astrocyte genes}
# Pull significant genes
astrocytes_sig_table <-
  astrocytes_marvel[["SJ.Gene.Cor"]][["Data"]]

# Pull significant genes
astrocytes_sig_genes <-
  astrocytes_marvel[["SJ.Gene.Cor"]][["Data"]]$gene_short_name

# Pull SJ locations
astrocytes_sig_coords <-
  astrocytes_marvel[["SJ.Gene.Cor"]][["Data"]]$coord.intron
```


```{r what are the significant astrocyte genes}
# Pull significant genes
astrocytes_sig_table <-
  astrocytes_marvel[["SJ.Gene.Cor"]][["Data"]]

# Pull significant genes
astrocytes_sig_genes <-
  astrocytes_marvel[["SJ.Gene.Cor"]][["Data"]]$gene_short_name

# Pull SJ locations
astrocytes_sig_coords <-
  astrocytes_marvel[["SJ.Gene.Cor"]][["Data"]]$coord.intron
```


```{r what are the significant astrocyte genes}
# Pull significant genes
oligodendrocytes_sig_table <-
  oligodendrocytes_marvel[["SJ.Gene.Cor"]][["Data"]]

# Pull significant genes
oligodendrocytes_sig_genes <-
  oligodendrocytes_marvel[["SJ.Gene.Cor"]][["Data"]]$gene_short_name

# Pull SJ locations
oligodendrocytes_sig_coords <-
  oligodendrocytes_marvel[["SJ.Gene.Cor"]][["Data"]]$coord.intron
```


```{r what are the significant astrocyte genes}
# Pull significant genes
opcs_sig_table <-
  opcs_marvel[["SJ.Gene.Cor"]][["Data"]]

# Pull significant genes
opcs_sig_genes <-
  opcs_marvel[["SJ.Gene.Cor"]][["Data"]]$gene_short_name

# Pull SJ locations
opcs_sig_coords <-
  opcs_marvel[["SJ.Gene.Cor"]][["Data"]]$coord.intron
```


```{r what are the significant astrocyte genes}
# Pull significant genes
vascular_sig_table <-
  vascular_marvel[["SJ.Gene.Cor"]][["Data"]]

# Pull significant genes
vascular_sig_genes <-
  vascular_marvel[["SJ.Gene.Cor"]][["Data"]]$gene_short_name

# Pull SJ locations
vascular_sig_coords <-
  vascular_marvel[["SJ.Gene.Cor"]][["Data"]]$coord.intron
```


```{r plot umaps for specific genes}
```

