---
title: "Calculate SJ expression by cell type"
author: "Emma Jones"
date: "2024-06-18"
output: html_document
---

The purpose of this script is to quantify the splice junction expression across cell types, both per cell and aggregated for each cell type. It creates a new supplemental table that is referred to in the manuscript. It is dependent on running all seurat scripts, and marvel scripts 01 through 04. Please use docker image setbp1_alternative_splicing:1.1.0. 

## Load packages

```{r load packages}
suppressPackageStartupMessages({
  # Load Lasseigne Lab standard packages
  library(here)
  library(styler)
  library(lintr)
  library(tidyverse)
  library(patchwork)
  library(ComplexHeatmap)
  library(viridis)
  library(cowplot)

  # Load MARVEL package
  library(MARVEL)

  # Load adjunct MARVEL packages
  library(ggnewscale)
  library(ggrepel)
  library(reshape2)
  library(plyr)
  library(stringr)
  library(textclean)
  library(AnnotationDbi)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(org.Mm.eg.db)
  library(gtools)
  library(GenomicRanges)
  library(IRanges)
  library(S4Vectors)
  library(wiggleplotr)
  library(Matrix)
  library(data.table)
  library(gridExtra)
  library(ggridges)
  library(GenomicFeatures)
  library(rtracklayer)
  library(ggpmisc)
})

# set seed
set.seed(123)

# source functions
source(here::here("src", "marvel", "functions.R"))
```

## Set colors

```{r set colors}
cell_type_colors <- c(
  `Astrocytes` = "#6CA9E2",
  `Excitatory Neurons` = "#98D070",
  `Inhibitory Neurons` = "#DEE971",
  `Microglia` = "#B898E4",
  `Oligodendrocytes` = "#4AD8E6",
  `OPCs` = "#0A9A8D",
  `Vascular Cells` = "#E28C67"
)
```

## Load data

First, we need to load in the splice junction usage data from script 04

```{r load in data}
# Load MARVEL data from marvel 04
load(here::here("data", "marvel", "marvel_04_split_counts.Rdata"))

# load in sj usage split by cell type
sj_usage_cell_type <- readRDS(here::here(
  "data", "marvel",
  "sj_usage_cell_type.rds"
))

# load in sj usage split by cell type and condition
sj_usage_condition <- readRDS(here::here(
  "data", "marvel",
  "sj_usage_condition.rds"
))

# Convert data type
sj_counts <- as(sj_counts, "CsparseMatrix")
gene_counts <- as(gene_counts, "CsparseMatrix")
```

# Get number of sjs detected per cell

```{r format metadata columns}
sample_metadata$num_sjs <- diff(sj_counts@p)
sample_metadata$num_genes <- diff(gene_counts@p)
sample_metadata$num_sjs_genes <- diff(sj_counts@p) / diff(gene_counts@p)
```

# do number of genes predict number of sjs?

```{r run linear regression}
# create linear model
sjs_genes_lm <- lm(num_sjs ~ num_genes, data = sample_metadata)

# get summary
summary(sjs_genes_lm)

summary(sjs_genes_lm)$r.squared
```


## Visualize genes, sjs, normalized sjs detected per cell

```{r genes per cell}
genes_per_cell_violin <- ggplot(
  sample_metadata,
  aes(
    x = cell_type, y = num_genes,
    fill = cell_type, alpha = 0.7
  )
) +
  geom_violin() +
  scale_fill_manual(values = cell_type_colors) +
  guides(fill = "none", alpha = "none") +
  geom_boxplot(width = 0.1) +
  theme_minimal(base_size = 14) +
  xlab("Cell Type") +
  ylab("Genes per Cell") +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold", color = "black"),
    axis.text.x = element_text(angle = 45, vjust = 0.95, hjust = 0.95)
  )

genes_per_cell_violin
```


```{r sjs per cell}
sjs_per_cell_violin <- ggplot(
  sample_metadata,
  aes(
    x = cell_type, y = num_sjs,
    fill = cell_type, alpha = 0.7
  )
) +
  geom_violin() +
  scale_fill_manual(values = cell_type_colors) +
  guides(fill = "none", alpha = "none") +
  geom_boxplot(width = 0.1) +
  theme_minimal(base_size = 14) +
  xlab("Cell Type") +
  ylab("SJs per Cell") +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold", color = "black"),
    axis.text.x = element_text(angle = 45, vjust = 0.95, hjust = 0.95)
  )

sjs_per_cell_violin
```


```{r sjs/genes per cell}
sjs_genes_per_cell_violin <- ggplot(
  sample_metadata,
  aes(
    x = cell_type, y = num_sjs_genes,
    fill = cell_type, alpha = 0.7
  )
) +
  geom_violin() +
  scale_fill_manual(values = cell_type_colors) +
  guides(fill = "none", alpha = "none") +
  geom_boxplot(width = 0.1) +
  theme_minimal(base_size = 14) +
  xlab("Cell Type") +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold", color = "black"),
    axis.text.x = element_text(angle = 45, vjust = 0.95, hjust = 0.95)
  ) +
  ylab("SJs per Cell/\nGenes per Cell")

sjs_genes_per_cell_violin
```
## Combine figure for supporting information figure

```{r cowplot and save}
supp_fig3 <- plot_grid(genes_per_cell_violin,
  sjs_per_cell_violin,
  sjs_genes_per_cell_violin,
  ncol = 1,
  labels = c("A", "B", "C")
)

supp_fig3


# save
png(here::here("results", "supplementary_figures", "supp_fig3.png"),
  width = 8, height = 12, units = "in", res = 300
)
supp_fig3
dev.off()
```

# Count total number of sjs detected in each cell type

```{r count sjs per cell type for each gene}
# Get all expressed gene names
all_expressed_genes <- rownames(gene_counts)[rowSums(gene_counts) > 0]

# Make NA matrix, with 7 columns for cell types
sj_per_gene <- matrix(NA, nrow = length(all_expressed_genes), ncol = 7)
# Add rownames
rownames(sj_per_gene) <- sort(all_expressed_genes)

# Initialize column
column <- 0

# Add in each table - these numbers are to exclude the functions
for (i in split_matrices_list) {
  column <- column + 1
  table <- get_sjs_per_gene(i)
  indices <- match(rownames(sj_per_gene), names(table))
  sj_per_gene[, column] <- ifelse(!is.na(indices), table[indices], 0)
}

# Add colnames
colnames(sj_per_gene) <-
  names(split_matrices_list)
```

Now save!

```{r save sjs per gene}
write.csv(sj_per_gene, here::here("results", "tables", "sj_per_gene.csv"))
```

#### Style

```{r tidy script}
style_file("05_cell_specific_sj_expr.Rmd")

lint("05_cell_specific_sj_expr.Rmd",
  linters = linters_with_defaults(
    object_length_linter = NULL,
    object_name_linter = NULL,
    object_usage_linter = NULL
  )
)
```

#### Software versions

My software versions will be commented below.

```{r versions}
sessionInfo()
```
R version 4.3.1 (2023-06-16)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3 
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so;  LAPACK version 3.10.0

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

time zone: Etc/UTC
tzcode source: system (glibc)

attached base packages:
[1] stats4    grid      stats     graphics  grDevices utils     datasets  methods  
[9] base     

other attached packages:
 [1] cowplot_1.1.1          ggpmisc_0.5.5          ggpp_0.5.6            
 [4] rtracklayer_1.60.1     GenomicFeatures_1.52.2 ggridges_0.5.4        
 [7] gridExtra_2.3          data.table_1.14.8      Matrix_1.6-1.1        
[10] wiggleplotr_1.24.0     GenomicRanges_1.52.1   GenomeInfoDb_1.36.4   
[13] gtools_3.9.4           org.Mm.eg.db_3.17.0    org.Hs.eg.db_3.17.0   
[16] clusterProfiler_4.8.3  AnnotationDbi_1.62.2   IRanges_2.34.1        
[19] S4Vectors_0.38.2       Biobase_2.60.0         BiocGenerics_0.46.0   
[22] textclean_0.9.3        plyr_1.8.9             reshape2_1.4.4        
[25] ggrepel_0.9.5          ggnewscale_0.4.9       MARVEL_2.0.5          
[28] viridis_0.6.4          viridisLite_0.4.2      ComplexHeatmap_2.16.0 
[31] patchwork_1.1.3        lubridate_1.9.3        forcats_1.0.0         
[34] stringr_1.5.1          dplyr_1.1.4            purrr_1.0.2           
[37] readr_2.1.4            tidyr_1.3.0            tibble_3.2.1          
[40] ggplot2_3.5.0          tidyverse_2.0.0        lintr_3.1.0           
[43] styler_1.10.2          here_1.0.1            

loaded via a namespace (and not attached):
  [1] splines_4.3.1               BiocIO_1.10.0              
  [3] filelock_1.0.2              bitops_1.0-7               
  [5] ggplotify_0.1.2             R.oo_1.25.0                
  [7] polyclip_1.10-6             XML_3.99-0.14              
  [9] rex_1.2.1                   lifecycle_1.0.4            
 [11] doParallel_1.0.17           rprojroot_2.0.3            
 [13] processx_3.8.2              lattice_0.21-8             
 [15] MASS_7.3-60                 backports_1.4.1            
 [17] magrittr_2.0.3              rmarkdown_2.25             
 [19] yaml_2.3.7                  remotes_2.4.2.1            
 [21] DBI_1.1.3                   RColorBrewer_1.1-3         
 [23] abind_1.4-5                 zlibbioc_1.46.0            
 [25] R.cache_0.16.0              R.utils_2.12.2             
 [27] ggraph_2.1.0                RCurl_1.98-1.12            
 [29] yulab.utils_0.1.0           rappdirs_0.3.3             
 [31] tweenr_2.0.2                circlize_0.4.15            
 [33] GenomeInfoDbData_1.2.10     enrichplot_1.20.3          
 [35] tidytree_0.4.5              qdapRegex_0.7.8            
 [37] MatrixModels_0.5-2          DelayedArray_0.26.7        
 [39] codetools_0.2-19            DOSE_3.26.2                
 [41] xml2_1.3.5                  ggforce_0.4.1              
 [43] tidyselect_1.2.1            shape_1.4.6                
 [45] aplot_0.2.2                 farver_2.1.1               
 [47] BiocFileCache_2.8.0         matrixStats_1.0.0          
 [49] GenomicAlignments_1.36.0    jsonlite_1.8.7             
 [51] GetoptLong_1.0.5            tidygraph_1.2.3            
 [53] survival_3.5-7              iterators_1.0.14           
 [55] foreach_1.5.2               tools_4.3.1                
 [57] progress_1.2.2              treeio_1.24.3              
 [59] Rcpp_1.0.12                 glue_1.7.0                 
 [61] xfun_0.40                   MatrixGenerics_1.12.3      
 [63] qvalue_2.32.0               withr_3.0.0                
 [65] fastmap_1.1.1               fansi_1.0.6                
 [67] SparseM_1.81                callr_3.7.3                
 [69] digest_0.6.33               timechange_0.2.0           
 [71] R6_2.5.1                    gridGraphics_0.5-1         
 [73] colorspace_2.1-0            GO.db_3.17.0               
 [75] biomaRt_2.56.1              RSQLite_2.3.1              
 [77] R.methodsS3_1.8.2           utf8_1.2.4                 
 [79] generics_0.1.3              S4Arrays_1.0.6             
 [81] prettyunits_1.2.0           graphlayouts_1.0.1         
 [83] httr_1.4.7                  scatterpie_0.2.1           
 [85] pkgconfig_2.0.3             gtable_0.3.4               
 [87] blob_1.2.4                  XVector_0.40.0             
 [89] shadowtext_0.1.2            htmltools_0.5.6.1          
 [91] fgsea_1.26.0                clue_0.3-65                
 [93] scales_1.3.0                cyclocomp_1.1.1            
 [95] png_0.1-8                   ggfun_0.1.3                
 [97] knitr_1.44                  rstudioapi_0.15.0          
 [99] tzdb_0.4.0                  rjson_0.2.21               
[101] curl_5.1.0                  nlme_3.1-163               
[103] cachem_1.0.8                GlobalOptions_0.1.2        
[105] parallel_4.3.1              HDO.db_0.99.1              
[107] restfulr_0.0.15             desc_1.4.2                 
[109] pillar_1.9.0                vctrs_0.6.5                
[111] dbplyr_2.3.4                cluster_2.1.4              
[113] evaluate_0.22               Rsamtools_2.16.0           
[115] cli_3.6.2                   compiler_4.3.1             
[117] rlang_1.1.3                 crayon_1.5.2               
[119] labeling_0.4.3              ps_1.7.5                   
[121] fs_1.6.3                    stringi_1.8.1              
[123] BiocParallel_1.34.2         munsell_0.5.1              
[125] Biostrings_2.68.1           lazyeval_0.2.2             
[127] quantreg_5.97               GOSemSim_2.26.1            
[129] hms_1.1.3                   bit64_4.0.5                
[131] KEGGREST_1.40.1             SummarizedExperiment_1.30.2
[133] igraph_1.5.1                memoise_2.0.1              
[135] ggtree_3.8.2                fastmatch_1.1-4            
[137] bit_4.0.5                   xmlparsedata_1.0.5         
[139] downloader_0.4              polynom_1.4-1              
[141] ape_5.7-1                   gson_0.1.0   
