---
title: "06_format_kidney_MARVEL_data"
author: "Tabea M. Soelter"
date: "2025-08-07"
output: html_document
---

MAKE SURE TO FIX THE SPLIT_SJ_INFO FUNCTION, BC IT DOES NOT WORK FOR KIDNEY. CURRENTLY SAVED IN 07_KIDNEY_MARVEL_TROUBLESHOOTING.R
- ALSO NEED TO FIX THE K6 FILE, SO MAKE SURE TO INCLUDE THE CODE IN THIS SCRIPT AS WELL DURING CLEAN-UP.


## Load Packages
```{r}
suppressPackageStartupMessages({
  library(here)
  library(styler)
  library(lintr)
  library(tidyverse)
  library(Seurat)
  library(patchwork)
  library(harmony)
  library(presto)
  library(MARVEL)
  library(ggnewscale)
  library(ggrepel)
  library(reshape2)
  library(plyr)
  library(textclean)
  library(AnnotationDbi)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(org.Mm.eg.db)
  library(gtools)
  library(GenomicRanges)
  library(IRanges)
  library(S4Vectors)
  library(wiggleplotr)
  library(Matrix)
  library(data.table)
  library(gridExtra)
  library(MatrixExtra)
  options("MatrixExtra.quick_show" = FALSE)
})

# set seed
set.seed(42)

# source functions
source(here::here("src", "marvel", "functions.R"))
```

## Import Expression Data
```{r}
annotated_kidney_samples <- read_rds(
  here::here("data", "seurat", "annotated_kidney_samples.rds")
)

gene_norm <- annotated_kidney_samples@assays[["RNA"]]@layers[["data"]]
gene_norm <- as(gene_norm, "TsparseMatrix")
gene_norm[gene_norm[, 1] != 0, ][1:5, 1:5]

gene_metadata <- annotated_kidney_samples@meta.data
gene_metadata <- tibble::rownames_to_column(gene_metadata, "cell.id")
head(gene_metadata)

gene_features <- rownames(annotated_kidney_samples)
head(gene_features)
gene_features <- as.data.frame(gene_features)
colnames(gene_features) <- "gene_short_name"

gene_counts <- annotated_kidney_samples@assays[["RNA"]]@layers[["counts"]]
gene_counts <- as(gene_counts, "TsparseMatrix")
gene_counts[gene_counts[, 1] != 0, ][1:5, 1:5]
```

## Import Splice Junction Data
```{r use function to extract splice information}
# sample K1
K1_sj_info <- split_sj_info(
  sample_id = "K1",
  condition = "WT"
)
# sample K2
K2_sj_info <- split_sj_info(
  sample_id = "K2",
  condition = "S858R"
)
# sample K3
K3_sj_info <- split_sj_info(
  sample_id = "K3",
  condition = "WT"
)
# sample K4
K4_sj_info <- split_sj_info(
  sample_id = "K4",
  condition = "S858R"
)
# sample K5
K5_sj_info <- split_sj_info(
  sample_id = "K5",
  condition = "WT"
)
# sample K6
K6_sj_info <- split_sj_info_ts(
  sample_id = "K6",
  condition = "S858R"
)
```

## Combine data
```{r}
# Make list of all matrices - make sure these are in the order of your gene data
matrices <- list(
  K2_sj_info, K4_sj_info, K6_sj_info, K1_sj_info, K3_sj_info,
  K5_sj_info
)

# Identify all unique features
all_features <- unique(unlist(lapply(matrices, rownames)))

# Align and combine matrices using custom function
combined_matrix <- do.call(cbind, lapply(matrices, align_matrix, all_features))

# Add correct colnames
colnames(combined_matrix) <- gene_metadata$cell.id

# Coerce to correct format
combined_matrix <- as(combined_matrix, "TsparseMatrix")

## NOT REALLY SURE WHAT THE PURPOSE OF THIS IS IN EJ'S CODE, I AM MOVING FORWARD WITHOUT IT.
# Check values where each new sample starts
#combined_matrix[combined_matrix[, 1] != 0, ][1:5, 1:5]

#K1_sj_info[K1_sj_info[, 1] != 0, ][1:5, 1:5]

# Check each sample - J13
#combined_matrix[combined_matrix[, 12297] != 0, ][1:5, 12297:12300]

#J13_sj_info[J13_sj_info[, 1] != 0, ][1:5, 1:9]

# Check each sample - J15
#combined_matrix[combined_matrix[, 14474] != 0, ][1:5, 14474:14482]

#J15_sj_info[J15_sj_info[, 1] != 0, ][1:5, 1:9]

# Check each sample - J2
#combined_matrix[combined_matrix[, 23427] != 0, ][1:5, 23427:23435]

#J2_sj_info[J2_sj_info[, 1] != 0, ][1:5, 1:9]

# Check each sample - J3
#combined_matrix[combined_matrix[, 30507] != 0, ][1:5, 30507:30515]

#J3_sj_info[J3_sj_info[, 1] != 0, ][1:5, 1:9]

# Check each sample - J4
#combined_matrix[combined_matrix[, 40978] != 0, ][1:5, 40978:40986]

#J4_sj_info[J4_sj_info[, 1] != 0, ][1:5, 1:9]

# rename matrix for ease
all_sj_counts <- combined_matrix

# get metadata (colnames)
sj_metadata <- colnames(combined_matrix)
sj_metadata <- as.data.frame(sj_metadata)
colnames(sj_metadata) <- "cell.id"

# get splice junction location names
sj_features <- rownames(combined_matrix)
sj_features <- as.data.frame(sj_features)
colnames(sj_features) <- "coord.intron"
```

### Get UMAP coordinates

```{r get UMAP coordinates}
umap_coords <-
  annotated_kidney_samples@reductions[["umap_harmony"]]@cell.embeddings

umap_coords <- tibble::rownames_to_column(as.data.frame(umap_coords), "cell.id")

colnames(umap_coords) <- c("cell.id", "x", "y")
```

### Get GTF

```{r get gtf in r env}
# genome directory
genome_dir <- "/data/project/lasseigne_lab/GENOME_dir/GENCODE_mm39/release_M31/"

# load GTF
m31_gtf <- as.data.frame(
  fread(paste0(genome_dir, "GTF/gencode.vM31.primary_assembly.annotation.gtf")),
  sep = "\t",
  header = FALSE,
  stringsAsFactors = FALSE)

# rename columns
colnames(m31_gtf) <- c("V1", "V2", "V3", "V4", "V5", "V6", "V7", "V8", "V9")
# drop chr from chromosome numbers in gtf column 1
m31_gtf$V1 <- substr(m31_gtf$V1, 4, nchar(m31_gtf$V1))
```

## Make MARVEL object

```{r make marvel object}
setbp1_marvel <- CreateMarvelObject.10x(
  gene.norm.matrix = gene_norm,
  gene.norm.pheno = gene_metadata,
  gene.norm.feature = gene_features,
  gene.count.matrix = gene_counts,
  gene.count.pheno = gene_metadata,
  gene.count.feature = gene_features,
  sj.count.matrix = all_sj_counts,
  sj.count.pheno = sj_metadata,
  sj.count.feature = sj_features,
  pca = umap_coords,
  gtf = m31_gtf
)
```

## Pre-process MARVEL object
```{r annotate gene metadta}
setbp1_marvel <- AnnotateGenes.10x(MarvelObject = setbp1_marvel)

head(setbp1_marvel$gene.metadata)

table(setbp1_marvel$gene.metadata$gene_type)
```

```{r annotate junction metadata}
setbp1_marvel <- AnnotateSJ.10x(MarvelObject = setbp1_marvel)

head(setbp1_marvel$sj.metadata)
```

Now, we want to validate these splice junctions whose start and end are mapped to the same gene (no multi-gene splicing is included downstream because of this step).

```{r validate splice junctions}
setbp1_marvel <- ValidateSJ.10x(MarvelObject = setbp1_marvel)
```

Lastly, we will remove gene that are non-coding genes.

```{r subset cds genes}
setbp1_marvel <- FilterGenes.10x(
  MarvelObject = setbp1_marvel,
  gene.type = "protein_coding"
)
```

Expected output:
18180 of 32530 genes met filtering criteria and retained
194653 of 205113 SJ met filtering criteria and retained
Normalised gene and SJ metadata and matrix updated

Last, before proceeding, we must check the alignment of the marvel object.

```{r check alignment}
setbp1_marvel <- CheckAlignment.10x(MarvelObject = setbp1_marvel)
```
Expected output:

Matching gene list in normalised and raw (count) gene matrices...
18180 genes found in normalised gene matrix
32568 genes found in raw (count) gene matrix
18180 overlapping genes found and retained
Matching cells across normalised gene, raw (count) gene, raw (count) SJ matrices...
75889 cells found in normalised gene matrix
75889 cells found in raw (count) gene matrix
75889 cells found in raw (count) SJ matrix
75889 overlapping cells found and retained
Checking column (sample) alignment...
Sample metadata and normalised gene matrix column names MATCHED
Normalised and raw (count) gene matrix column names MATCHED
Raw (Count) gene and SJ matrix column names MATCHED
Checking row (gene names/SJ coordinates) alignment...
Gene metadata and normalised gene matrix row names MATCHED
Normalised and raw (count) gene matrix row names MATCHED
SJ metadata and raw (count) SJ matrix row names MATCHED
75889 cells and 15561 genes consisting of 194653 splice junctions included for further analysis

Looks good! Now we can save our marvel object before proceeding.

```{r save marvel object}
# save object
write_rds(setbp1_marvel,
  file = here::here("data", "marvel", "setbp1_kidney_marvel_aligned.rds")
)
```

