---
title: "Annotate Cell Types in Seurat"
author: "Emma Jones"
date: "2023-11-15"
output: html_document
---

The purpose of this script is to annotate cell types using seurat. It is dependent on script 01. Please use docker image setbp1_alternative_splicing:1.0.4

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Packages

```{r load packages}
suppressPackageStartupMessages({
  # Load Lasseigne Lab standard packages
  library(here)
  library(styler)
  library(lintr)

  # Load Seurat and other single-cell packages
  library(tidyverse)
  library(Seurat)
  library(patchwork)
  library(harmony)
  library(presto)
})

# set seed
set.seed(123)

# source functions
source(here("src", "functions.R"))
```

## Load data

```{r load in data}
clustered_brain_samples <- readRDS(here(
  "data", "seurat",
  "clustered_brain_samples.rds"
))
```

## Find marker genes

```{r find markers}
# set identity to best clustering after exploring data
clustered_brain_samples <- SetIdent(clustered_brain_samples,
  value = "RNA_snn_res.0.75"
)

DimPlot(clustered_brain_samples, label = TRUE)

marker_genes <- FindAllMarkers(clustered_brain_samples,
  logfc.threshold = 0.2,
  assay = "RNA",
  min.pct = 0.01,
  only.pos = TRUE
)

top10 <- marker_genes %>%
  group_by(cluster) %>%
  top_n(-10, p_val_adj)
```

## Use canonical cell type markers

### Neuron cell annotation

The markers I am using for neurons are:

All neurons - Snap25
Excitatory Neurons - Slc17a7
Inhibitory Neurons - Gad1, Gad2
We also looked at markers for specific kinds of inhibitory neurons: Sst, Pvalb, and Vip

```{r featureplots for neurons}
FeaturePlot(clustered_brain_samples,
  features = c("Snap25"),
  label = TRUE, repel = TRUE
) # clusters 7, 20, 3, 14, 29, 31 seem to have no expression

FeaturePlot(clustered_brain_samples,
  features = c("Slc17a7"),
  label = TRUE, repel = TRUE
) # excitatory neurons
FeaturePlot(clustered_brain_samples,
  features = c("Pcp4"),
  label = TRUE
)

FeaturePlot(clustered_brain_samples,
  features = c("Gad1"),
  label = TRUE, repel = TRUE
) # more inhibitory neurons - 11, 12, 9, 18
FeaturePlot(clustered_brain_samples,
  features = c("Gad2"),
  label = TRUE
)

FeaturePlot(clustered_brain_samples,
  features = c("Sst"),
  label = TRUE
) # sst inhibitory neurons - 12
FeaturePlot(clustered_brain_samples,
  features = c("Pvalb"),
  label = TRUE
) # pvalb inhibitory neurons - 9
FeaturePlot(clustered_brain_samples,
  features = c("Vip"),
  label = TRUE
) # vip - 11
```

### Glia cell annotation

The markers I used for non-neuronal cells:
Microglia - Cx3cr1, Cd68
Astrocytes - Slc1a3, Gja1
Vascular Cells (Pericytes/Fibroblasts/VLMCs/Endothelial Cells) - Vtn, Mgp, Bnc2, Pdgfrb, Col1a1, Dcn, Flt1
OPCs - Pdgfra
Oligodendrocytes - Mbp

```{r more exploratory cluster annotation non-neuronal cells}
FeaturePlot(clustered_brain_samples,
  features = c("Cx3cr1"),
  label = TRUE
) # microglia - 14
FeaturePlot(clustered_brain_samples,
  features = c("Cd68"),
  label = TRUE, repel = TRUE
) # cluster 14 (macrophage and microglia)

FeaturePlot(clustered_brain_samples,
  features = c("Slc1a3", "Gja1"),
  label = TRUE
) # astrocyte marker - 3

FeaturePlot(clustered_brain_samples,
  features = c("Mbp"),
  label = TRUE, repel = TRUE
) # oligodendrocytes are 7, 20 ,22, 23, 31

FeaturePlot(clustered_brain_samples,
  features = c("Pdgfra"),
  label = TRUE
) # cluster 13 is OPCs

FeaturePlot(clustered_brain_samples,
  features = c("Col1a1"),
  label = TRUE, repel = TRUE
) # fibroblast marker

FeaturePlot(clustered_brain_samples,
  features = c("Dcn"),
  label = TRUE, repel = TRUE
) # fibroblast marker

FeaturePlot(clustered_brain_samples,
  features = c("Mgp"),
  label = TRUE, repel = TRUE
) # fibroblast and vascular cell marker

FeaturePlot(clustered_brain_samples,
  features = c("Cald1"),
  label = TRUE, repel = TRUE
) # fibroblast marker - 26 and 32

FeaturePlot(clustered_brain_samples,
  features = c("Vtn"),
  label = TRUE, repel = TRUE
) # pericyte marker - 26

FeaturePlot(clustered_brain_samples,
  features = c("Pdgfrb"),
  label = TRUE, repel = TRUE
) # pericyte marker - 26

FeaturePlot(clustered_brain_samples,
  features = c("Flt1"),
  label = TRUE, repel = TRUE
) # endothelial cells - 29

FeaturePlot(clustered_brain_samples,
  features = c("Cldn5"),
  label = TRUE, repel = TRUE
) # endothelial cells - 29

FeaturePlot(clustered_brain_samples,
  features = c("Pecam1"),
  label = TRUE, repel = TRUE
) # endothelial cell marker
```

### Assign cell types

Next, we assign cell types.

```{r assign cell types}
annotated_brain_samples <- RenameIdents(clustered_brain_samples,
  `1` = "Excitatory Neurons",
  `2` = "Excitatory Neurons",
  `3` = "Astrocytes",
  `4` = "Excitatory Neurons",
  `5` = "Excitatory Neurons",
  `6` = "Excitatory Neurons",
  `7` = "Oligodendrocytes",
  `8` = "Excitatory Neurons",
  `9` = "Inhibitory Neurons",
  `10` = "Excitatory Neurons",
  `11` = "Inhibitory Neurons",
  `12` = "Inhibitory Neurons",
  `13` = "OPCs",
  `14` = "Microglia",
  `15` = "Excitatory Neurons",
  `16` = "Excitatory Neurons",
  `17` = "Excitatory Neurons",
  `18` = "Inhibitory Neurons",
  `19` = "Excitatory Neurons",
  `20` = "Oligodendrocytes",
  `21` = "Excitatory Neurons",
  `22` = "Test",
  `23` = "Oligodendrocytes",
  `24` = "Excitatory Neurons",
  `25` = "Excitatory Neurons",
  `26` = "Vascular Cells",
  `27` = "Excitatory Neurons",
  `28` = "Excitatory Neurons",
  `29` = "Vascular Cells",
  `30` = "Excitatory Neurons",
  `31` = "Oligodendrocytes",
  `32` = "Vascular Cells"
)

annotated_brain_samples <-
  AddMetaData(
    object = annotated_brain_samples,
    as.vector(annotated_brain_samples@active.ident),
    col.name = "cell_type"
  )
```

Now that we have cell types identified, we can see what the final UMAP looks like.

```{r plot umap}
DimPlot(annotated_brain_samples)
```

There is a bit of oligodendrocytes that dont express Mbp and were not easily distinguishable as cluster 22. So, I want to subcluster 22.

### Subcluster 22

We can use Seurat's FindSubCluster function to help here

```{r subcluster 22}
subclustered_brain_samples <- FindSubCluster(
  clustered_brain_samples,
  22,
  "RNA_snn",
  subcluster.name = "subcluster_22",
  resolution = 0.05,
  algorithm = 4
)

DimPlot(subclustered_brain_samples, group.by = "subcluster_22")
```

Reannotate based on the featureplots of these subclustered cells.

```{r look at astrocyte and oligodendrocyte markers}
subclustered_brain_samples <- SetIdent(subclustered_brain_samples,
  value = "subcluster_22"
)

FeaturePlot(subclustered_brain_samples,
  features = c("Slc1a3", "Gja1"),
  label = TRUE
) # astrocyte marker - 3

FeaturePlot(subclustered_brain_samples,
  features = c("Mbp"),
  label = TRUE, repel = TRUE
) # oligodendrocyte marker
```

Set new names for new cell type annotation.

```{r annotate subclusters}
annotated_brain_samples <- RenameIdents(subclustered_brain_samples,
  `1` = "Excitatory Neurons",
  `2` = "Excitatory Neurons",
  `3` = "Astrocytes",
  `4` = "Excitatory Neurons",
  `5` = "Excitatory Neurons",
  `6` = "Excitatory Neurons",
  `7` = "Oligodendrocytes",
  `8` = "Excitatory Neurons",
  `9` = "Inhibitory Neurons",
  `10` = "Excitatory Neurons",
  `11` = "Inhibitory Neurons",
  `12` = "Inhibitory Neurons",
  `13` = "OPCs",
  `14` = "Microglia",
  `15` = "Excitatory Neurons",
  `16` = "Excitatory Neurons",
  `17` = "Excitatory Neurons",
  `18` = "Inhibitory Neurons",
  `19` = "Excitatory Neurons",
  `20` = "Oligodendrocytes",
  `21` = "Excitatory Neurons",
  `22_1` = "Astrocytes",
  `22_2` = "Oligodendrocytes",
  `22_3` = "Oligodendrocytes",
  `23` = "Oligodendrocytes",
  `24` = "Excitatory Neurons",
  `25` = "Excitatory Neurons",
  `26` = "Vascular Cells",
  `27` = "Excitatory Neurons",
  `28` = "Excitatory Neurons",
  `29` = "Vascular Cells",
  `30` = "Excitatory Neurons",
  `31` = "Oligodendrocytes",
  `32` = "Vascular Cells"
)

annotated_brain_samples <-
  AddMetaData(
    object = annotated_brain_samples,
    as.vector(annotated_brain_samples@active.ident),
    col.name = "cell_type"
  )
```

### Visualize finished UMAP

Let's look at the new umap.

```{r plot umap}
DimPlot(annotated_brain_samples)
```

Yes, I think this is better.

### Save

Now, we can save the umap for later.

```{r save umap}
png(here("results", "seurat_outputs", "annotated_umap.png"))
DimPlot(annotated_brain_samples)
dev.off()
```

Save annotated object!

```{r save annotated object}
# save
saveRDS(annotated_brain_samples,
  file = here("data", "seurat", "annotated_brain_samples.rds")
)
```

#### Style

```{r tidy script}
style_file("02_annotate_cell_types.Rmd")

lint("02_annotate_cell_types.Rmd",
  linters = linters_with_defaults(
    object_length_linter = NULL,
    object_name_linter = NULL,
    object_usage_linter = NULL
  )
)
```

#### Software versions

My software versions will be commented below.

```{r versions}
sessionInfo()
```
R version 4.3.1 (2023-06-16)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3 
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so;  LAPACK version 3.10.0

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

time zone: Etc/UTC
tzcode source: system (glibc)

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] presto_1.0.0       data.table_1.14.8  harmony_1.1.0      Rcpp_1.0.11       
 [5] patchwork_1.1.3    Seurat_5.0.0       SeuratObject_5.0.0 sp_2.1-1          
 [9] lubridate_1.9.3    forcats_1.0.0      stringr_1.5.1      dplyr_1.1.3       
[13] purrr_1.0.2        readr_2.1.4        tidyr_1.3.0        tibble_3.2.1      
[17] ggplot2_3.4.4      tidyverse_2.0.0    lintr_3.1.0        styler_1.10.2     
[21] here_1.0.1        

loaded via a namespace (and not attached):
  [1] RColorBrewer_1.1-3     rstudioapi_0.15.0      jsonlite_1.8.7        
  [4] magrittr_2.0.3         spatstat.utils_3.0-4   farver_2.1.1          
  [7] rmarkdown_2.25         vctrs_0.6.4            ROCR_1.0-11           
 [10] spatstat.explore_3.2-5 htmltools_0.5.6.1      sctransform_0.4.1     
 [13] parallelly_1.36.0      KernSmooth_2.23-22     htmlwidgets_1.6.2     
 [16] desc_1.4.2             ica_1.0-3              plyr_1.8.9            
 [19] plotly_4.10.3          zoo_1.8-12             igraph_1.5.1          
 [22] mime_0.12              lifecycle_1.0.4        pkgconfig_2.0.3       
 [25] Matrix_1.6-1.1         R6_2.5.1               fastmap_1.1.1         
 [28] fitdistrplus_1.1-11    future_1.33.0          shiny_1.7.5.1         
 [31] digest_0.6.33          colorspace_2.1-0       ps_1.7.5              
 [34] tensor_1.5             rprojroot_2.0.3        RSpectra_0.16-1       
 [37] irlba_2.3.5.1          labeling_0.4.3         progressr_0.14.0      
 [40] fansi_1.0.5            spatstat.sparse_3.0-3  timechange_0.2.0      
 [43] polyclip_1.10-6        abind_1.4-5            httr_1.4.7            
 [46] compiler_4.3.1         remotes_2.4.2.1        withr_2.5.2           
 [49] backports_1.4.1        fastDummies_1.7.3      R.utils_2.12.2        
 [52] MASS_7.3-60            tools_4.3.1            lmtest_0.9-40         
 [55] httpuv_1.6.12          future.apply_1.11.0    goftest_1.2-3         
 [58] R.oo_1.25.0            glue_1.6.2             callr_3.7.3           
 [61] nlme_3.1-163           R.cache_0.16.0         promises_1.2.1        
 [64] grid_4.3.1             Rtsne_0.16             cluster_2.1.4         
 [67] reshape2_1.4.4         generics_0.1.3         gtable_0.3.4          
 [70] spatstat.data_3.0-3    tzdb_0.4.0             R.methodsS3_1.8.2     
 [73] hms_1.1.3              xml2_1.3.5             utf8_1.2.4            
 [76] spatstat.geom_3.2-7    RcppAnnoy_0.0.21       ggrepel_0.9.4         
 [79] RANN_2.6.1             pillar_1.9.0           spam_2.10-0           
 [82] RcppHNSW_0.5.0         later_1.3.1            splines_4.3.1         
 [85] lattice_0.21-8         deldir_1.0-9           survival_3.5-7        
 [88] tidyselect_1.2.0       miniUI_0.1.1.1         pbapply_1.7-2         
 [91] knitr_1.44             gridExtra_2.3          scattermore_1.2       
 [94] xfun_0.40              matrixStats_1.0.0      rex_1.2.1             
 [97] stringi_1.8.1          lazyeval_0.2.2         yaml_2.3.7            
[100] evaluate_0.22          codetools_0.2-19       cli_3.6.1             
[103] uwot_0.1.16            xtable_1.8-4           reticulate_1.34.0     
[106] munsell_0.5.0          processx_3.8.2         xmlparsedata_1.0.5    
[109] spatstat.random_3.2-1  globals_0.16.2         png_0.1-8             
[112] parallel_4.3.1         ellipsis_0.3.2         dotCall64_1.1-0       
[115] listenv_0.9.0          viridisLite_0.4.2      cyclocomp_1.1.1       
[118] scales_1.2.1           ggridges_0.5.4         leiden_0.4.3          
[121] crayon_1.5.2           rlang_1.1.2            cowplot_1.1.1    


## Archive
Look at cortical layer markers - I am leaving this code here in case we can to subcluster excitatory neurons later.

```{r cortical layer markers}
# layer 2/3
FeaturePlot(clustered_brain_samples, features = c("Lamp5"), label = TRUE)
# layer 2/4
FeaturePlot(clustered_brain_samples, features = c("Cux2"), label = TRUE)
# layer 2/5
FeaturePlot(clustered_brain_samples, features = c("Rasgrf2"), label = TRUE)
# layer 4/5
FeaturePlot(clustered_brain_samples, features = c("Rorb"), label = TRUE)
# layer 5
FeaturePlot(clustered_brain_samples,
  features = c("Kcnk2", "Sulf2", "Pcp4", "Fezf2"), label = TRUE
)
# layer 5/6
FeaturePlot(clustered_brain_samples,
  features = c("Etv1", "Rprm", "Rxfp1", "Foxp2"), label = TRUE
)
# layer 6
FeaturePlot(clustered_brain_samples,
  features = c("Syt6", "Oprk1", "Nr4a2", "Synpr"), label = TRUE
)
```
