---
title: "Create Figure 3"
author: "Emma Jones"
date: "2024-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Create figure 3

The purpose of this script is to make and complete figure 3. This figure has an upset plot of all alternatively spliced genes, and functional enrichment analysis of cell-type specific AS genes.

## Load packages

```{r load packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(styler)
  library(lintr)
  library(cowplot)
  library(patchwork)
  library(viridis)
  library(org.Mm.eg.db)
  library(gprofiler2)
  library(AnnotationDbi)
  library(clusterProfiler)
  library(ComplexHeatmap)
  library(readxl)
})
```


```{r load marvel object and significant gene tables}
# Load MARVEL object
setbp1_marvel <- read_rds(here::here(
  "data", "marvel",
  "setbp1_marvel_aligned.rds"
))

# Retrieve sample metadata
sample_metadata <- setbp1_marvel$sample.metadata

# Load significant gene tables
load(here::here("data", "marvel", "significant_tables.RData"))

# Pull cell types and matching ids
cell_group_list <- sample_metadata %>%
  group_split(cell_type, .keep = TRUE) %>%
  map(~ set_names(.$cell.id, .$cell_type[1]))

# Rename
cell_group_list <- set_names(cell_group_list, c(
  "Astrocytes", "Excitatory Neurons",
  "Inhibitory Neurons", "Microglia", "OPCs",
  "Oligodendrocytes", "Vascular Cells"
))
```

### Prep data for UpSet Plot with ComplexHeatmap

```{r make upset plot}
# make list of sig tables
tables_list <- mget(ls(pattern = "_sig_table"))

# remove all cell types from list
tables_list <- tables_list[2:8]

# pull out only genes
gene_lists <- map(tables_list, ~ select(.x, gene_short_name))

# make into character list
gene_lists <- map(gene_lists, ~ flatten(.))

# create matrix
gene_lists_mat <- ComplexHeatmap::list_to_matrix(gene_lists)

# rename columns
colnames(gene_lists_mat) <- c(
  "Astrocytes", "Excitatory Neurons",
  "Inhibitory Neurons", "Microglia",
  "Oligodendrocytes", "OPCs", "Vascular Cells"
)

# generate combination matrix
gene_lists_comb_mat <- make_comb_mat(gene_lists_mat)
```

### Draw and save UpSet Plot

```{r draw and save upsetplot}
gene_lists_upset_plot <- UpSet(gene_lists_comb_mat,
  comb_col = viridis(8)[2:7][comb_degree(gene_lists_comb_mat)],
  top_annotation = upset_top_annotation(gene_lists_comb_mat,
    add_numbers = TRUE, annotation_name_gp = gpar(fontface = "bold"),
    axis_param = list(gp = gpar(fontface = "bold")),
    numbers_gp = gpar(fontface = "bold")
  ),
  right_annotation = upset_right_annotation(gene_lists_comb_mat,
    add_numbers = TRUE, annotation_name_gp = gpar(fontface = "bold"),
    axis_param = list(gp = gpar(fontface = "bold")),
    numbers_gp = gpar(fontface = "bold")
  ),
  row_names_gp = gpar(fontface = "bold")
)

# view upsetplot
gene_lists_upset_plot

# convert to grob
gene_lists_upset_plot <- grid.grabExpr(draw(gene_lists_upset_plot))
```

There is one gene that has significant change in splice junction usage in all non-vascular cell types. This is the Son gene.

## FEA for each cell type

```{r get background genes}
background_genes <- setbp1_marvel[["gene.count.matrix"]]@Dimnames[[1]]
```

To avoid aggregating cells and losing specific signals, lets only include genes from each cell type.

```{r run fea on each cell type}
cell_type_goresults <- lapply(gene_lists, function(x) {
  neuron_gostres <- gost(
    query = unlist(x),
    organism = "mmusculus", ordered_query = FALSE,
    multi_query = FALSE, significant = TRUE, exclude_iea = FALSE,
    measure_underrepresentation = FALSE, evcodes = TRUE,
    user_threshold = 0.05, correction_method = "g_SCS",
    domain_scope = "custom", custom_bg = background_genes,
    numeric_ns = "", sources = NULL, as_short_link = FALSE
  )
})
```

```{r save and plot new go results}
# save rds
write_rds(
  cell_type_goresults,
  here::here("data", "marvel", "cell_type_goresults.rds")
)
```

```{r plot oligodendrocytes and excitatory neuron terms}
excitatory_panel <- ggplot(
  cell_type_goresults[["excitatory_sig_table"]][["result"]],
  aes(
    x = precision,
    y = reorder(term_name, -p_value),
    size = intersection_size,
    color = p_value
  )
) +
  geom_point() +
  scale_color_viridis(direction = 1, option = "mako") +
  theme_minimal(base_size = 12) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold", color = "black"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    title = element_text(face = "bold")
  ) +
  xlab("Precision") +
  ylab("Term Name") +
  labs(color = "p-value", size = "Intersection Size") +
  ggtitle("Excitatory Neuron Functional\nEnrichment Analysis") +
  guides(
    size = guide_legend(order = 1),
    color = guide_colorbar(order = 2)
  )

excitatory_panel

oligo_panel <- ggplot(
  cell_type_goresults[["oligodendrocytes_sig_table"]][["result"]],
  aes(
    x = precision,
    y = reorder(term_name, -p_value),
    size = intersection_size,
    color = p_value
  )
) +
  geom_point() +
  scale_color_viridis(direction = 1, option = "mako") +
  theme_minimal(base_size = 12) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold", color = "black"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    title = element_text(face = "bold")
  ) +
  xlab("Precision") +
  ylab("Term Name") +
  labs(color = "p-value", size = "Intersection Size") +
  ggtitle("Oligodendrocyte Functional\nEnrichment Analysis") +
  guides(
    size = guide_legend(order = 1),
    color = guide_colorbar(order = 2)
  )

oligo_panel
```

```{r arrange all results}
bottom_plots <- plot_grid(oligo_panel, excitatory_panel,
  labels = c("B", "C"), nrow = 1
)

figure_3 <- plot_grid(gene_lists_upset_plot, bottom_plots,
  ncol = 1,
  labels = c("A", "")
)

figure_3
```


```{r save figure 3}
# save
png(here::here("results", "final_outputs", "figure3.png"),
  width = 12, height = 8, units = "in", res = 300
)
figure_3
dev.off()
```

```{r make supp information for other cell types}
inhibitory_panel <- ggplot(
  cell_type_goresults[["inhibitory_sig_table"]][["result"]],
  aes(
    x = precision,
    y = reorder(term_name, -p_value),
    size = intersection_size,
    color = p_value
  )
) +
  geom_point() +
  scale_color_viridis(direction = 1, option = "mako") +
  theme_minimal(base_size = 12) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold", color = "black"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    title = element_text(face = "bold")
  ) +
  xlab("Precision") +
  ylab("Term Name") +
  labs(color = "p-value", size = "Intersection Size") +
  ggtitle("Inhibitory Neuron Functional\nEnrichment Analysis") +
  guides(
    size = guide_legend(order = 1),
    color = guide_colorbar(order = 2)
  )

inhibitory_panel

astro_panel <- ggplot(
  cell_type_goresults[["astrocytes_sig_table"]][["result"]],
  aes(
    x = precision,
    y = reorder(term_name, -p_value),
    size = intersection_size,
    color = p_value
  )
) +
  geom_point() +
  scale_color_viridis(direction = 1, option = "mako") +
  theme_minimal(base_size = 12) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold", color = "black"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    title = element_text(face = "bold")
  ) +
  xlab("Precision") +
  ylab("Term Name") +
  labs(color = "p-value", size = "Intersection Size") +
  ggtitle("Astrocyte Functional\nEnrichment Analysis") +
  guides(
    size = guide_legend(order = 1),
    color = guide_colorbar(order = 2)
  )

astro_panel

vascular_panel <- ggplot(
  cell_type_goresults[["vascular_sig_table"]][["result"]],
  aes(
    x = precision,
    y = reorder(term_name, -p_value),
    size = intersection_size,
    color = p_value
  )
) +
  geom_point() +
  scale_color_viridis(direction = 1, option = "mako") +
  theme_minimal(base_size = 12) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold", color = "black"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    title = element_text(face = "bold")
  ) +
  xlab("Precision") +
  ylab("Term Name") +
  labs(color = "p-value", size = "Intersection Size") +
  ggtitle("Vascular Cell Functional\nEnrichment Analysis") +
  guides(
    size = guide_legend(order = 1),
    color = guide_colorbar(order = 2)
  )

vascular_panel

opc_panel <- ggplot(
  cell_type_goresults[["opcs_sig_table"]][["result"]],
  aes(
    x = precision,
    y = reorder(term_name, -p_value),
    size = intersection_size,
    color = p_value
  )
) +
  geom_point() +
  scale_color_viridis(direction = 1, option = "mako") +
  theme_minimal(base_size = 12) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold", color = "black"),
    legend.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    title = element_text(face = "bold")
  ) +
  xlab("Precision") +
  ylab("Term Name") +
  labs(color = "p-value", size = "Intersection Size") +
  ggtitle("OPC Functional\nEnrichment Analysis") +
  guides(
    size = guide_legend(order = 1),
    color = guide_colorbar(order = 2)
  )

opc_panel
```

```{r format and save}
supp_4 <- plot_grid(inhibitory_panel, astro_panel, opc_panel, vascular_panel,
  labels = c("A", "B", "C", "D")
)

supp_4

# save
png(here::here("results", "final_outputs", "fea_other_cell_types.png"),
  width = 11, height = 8, units = "in", res = 300
)
supp_4
dev.off()
```


## Compare my 34 genes to 38 splicing genes in blood

This supplementary table was downloaded from https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9490002/ 

```{r compare to splicing genes in blood}
blood_aberrantly_spliced <- read_xlsx(
  path = here("doc", "Data_Sheet_2.xlsx"),
  sheet = 2, skip = 1
)

# there is a mistake in their table and unknown genes all have the same gene
# symbol (KNOWN) even though they have different locations

# the authors of the paper count this as 38 genes, when it is only 37
blood_aberrantly_spliced$hgncSymbol[
  blood_aberrantly_spliced$hgncSymbol == "KNOWN"
] <- NA

# drop na
blood_aberrantly_spliced <- drop_na(blood_aberrantly_spliced)

blood_aberrantly_spliced_genes <- unique(blood_aberrantly_spliced$hgncSymbol)

# convert human to mouse
source(here("src", "seurat", "functions.R"))
mouse_blood_spliced_genes <-
  convert_human_gene_list(blood_aberrantly_spliced_genes)

# get sig tables
matching_dfs <- mget(ls(pattern = "sig_table"))

# get the column you want
as_genes <- unique(unlist(lapply(matching_dfs, function(df_name) {
  df_name$gene_short_name
})))

# look at overlap
overlapping_genes <- intersect(mouse_blood_spliced_genes, as_genes)

# no overlap
```

#### Style

```{r tidy script}
style_file("figure_3.Rmd")

lint("figure_3.Rmd",
  linters = linters_with_defaults(
    object_length_linter = NULL,
    object_name_linter = NULL,
    object_usage_linter = NULL
  )
)
```

#### Software versions

My software versions will be commented below.

```{r versions}
sessionInfo()
```
