---
title: "Create Supp Figure 1"
author: "Emma Jones"
output: html_document
date: "2024-02-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The purpose of this script is to create a finalized version of supporting information figure 2. It is dependent on Seurat scripts 01 and 02. Run in docker 1.0.6.

### Load Packages

```{r load packages}
suppressPackageStartupMessages({
  # Load Lasseigne Lab standard packages
  library(here)
  library(styler)
  library(lintr)

  # Load Seurat and other single-cell packages
  library(tidyverse)
  library(Seurat)
  library(patchwork)
  library(harmony)
  library(presto)
  library(cowplot)
})
```

### Load data

We need to load in the annotated Seurat object.

```{r load data}
# load in data
annotated_brain_samples <-
  read_rds(file = here("data", "seurat", "annotated_brain_samples.rds"))
```

### Set colors

Set color palette for paper. These colors are colorblind-friendly and will be used throughout the paper.

```{r set colors}
cell_type_colors <- c(
  `Astrocytes` = "#6CA9E2",
  `Excitatory Neurons` = "#98D070",
  `Inhibitory Neurons` = "#DEE971",
  `Microglia` = "#B898E4",
  `Oligodendrocytes` = "#4AD8E6",
  `OPCs` = "#0A9A8D",
  `Vascular Cells` = "#E28C67"
)

condition_colors <- c(
  `mutant` = "#666666",
  `wildtype` = "#C1C1C1"
)

other_colors <- c("#0346a3", "#076466", "#2a6607")
```

### Panel A

I am first going to create a UMAP showing cell types.

```{r plot UMAP of cell types}
# make umap to show cell types
umap_1 <- DimPlot(annotated_brain_samples,
  cols = cell_type_colors,
  label = TRUE,
  label.box = TRUE,
  label.color = "white",
  label.size = 3.75,
  repel = TRUE,
  seed = 123
) +
  labs(color = "Cell Type") +
  theme(
    axis.text = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  ) +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  NoLegend()

# show umap
umap_1
```

### Panel B

Panel B will be a dot plot of cell type markers.

```{r plot dotplot of cell types}
# create dotplot
dotplot <- DotPlot(annotated_brain_samples,
  cols = c("lightgray", other_colors[2]),
  features = c(
    "Vtn", "Cx3cr1", "Pdgfra", "Gad2", "Mbp",
    "Aqp4", "Slc17a7"
  )
) +
  theme(
    legend.title = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  ylab("Cell Type")


# display dotplot
dotplot
```

### Panel C

Panel C will be a UMAP showing condition integration.

```{r plot UMAP of conditions}
# make umap for mutants and wildtype
umap_2 <- DimPlot(annotated_brain_samples,
  group.by = "seq_folder",
  shuffle = TRUE,
  seed = 123
) +
  theme(
    legend.title = element_text(face = "bold"),
    legend.text = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  ) +
  scale_color_manual(
    labels = c("S858R+/-", "Wildtype"),
    values = condition_colors
  ) +
  labs(color = "Condition", title = NULL) +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  NoLegend()

# show umap
umap_2
```
### Panel D

Make panel D, which will be bar plots taht show cell type proportions.

```{r show cell type proportions}
# aggregate cell type information
df <- table(
  Idents(annotated_brain_samples),
  annotated_brain_samples$seq_folder
) %>% as.data.frame()

# change classes for plotting
df$Var1 <- as.character(df$Var1)

# rename columns
colnames(df) <- c("Cell_Type", "Condition", "Freq")

# make silly data for legend
df3 <- data.frame(
  "Cell_Type" = as.numeric(c(NA, NA)),
  "Condition" = c("S858R+/-", "Wildtype"),
  "Freq" = as.numeric(c(NA, NA))
)

# make actual barplot
barplot <- ggplot(data = df, aes(
  x = fct_rev(Cell_Type),
  y = Freq, fill = Condition
)) +
  theme_minimal(base_size = 15) +
  geom_col(position = "fill", width = 0.75, show.legend = FALSE) +
  scale_fill_manual(
    labels = c("S858R+/-", "Wildtype"),
    values = condition_colors
  ) +
  theme(
    legend.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold", color = "black"),
    legend.text = element_text(face = "bold", size = 12),
    axis.text.x = element_text()
  ) +
  coord_flip() +
  xlab("Cell Type") +
  ylab("Proportion") +
  geom_point(data = df3, aes(
    x = Cell_Type,
    y = Freq, color = Condition
  ), size = 4) +
  scale_color_manual(
    labels = c("S858R+/-", "Wildtype"),
    values = c("#666666", "#C1C1C1")
  ) +
  theme(legend.key = element_blank())

# display barplot
barplot
```

## Save final figure

Now, we can pull everything together.

```{r plot everything together and save}
figure_1 <- plot_grid(umap_1, dotplot, umap_2, barplot,
  labels = c("A", "B", "C", "D")
)

png(here::here("results", "supplementary_figures", "supp_figure1.png"),
  width = 12, height = 8, units = "in", res = 300
)
figure_1
dev.off()
```

#### Style

```{r tidy script}
style_file("supp_figure_1.Rmd")

lint("supp_figure_1.Rmd",
  linters = linters_with_defaults(
    object_length_linter = NULL,
    object_name_linter = NULL,
    object_usage_linter = NULL
  )
)
```

#### Software versions

My software versions will be commented below.

```{r versions}
sessionInfo()
```
R version 4.3.1 (2023-06-16)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3 
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so;  LAPACK version 3.10.0

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

time zone: Etc/UTC
tzcode source: system (glibc)

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] cowplot_1.1.1      presto_1.0.0       data.table_1.14.8  harmony_1.1.0     
 [5] Rcpp_1.0.11        patchwork_1.1.3    Seurat_5.0.0       SeuratObject_5.0.0
 [9] sp_2.1-1           lubridate_1.9.3    forcats_1.0.0      stringr_1.5.1     
[13] dplyr_1.1.3        purrr_1.0.2        readr_2.1.4        tidyr_1.3.0       
[17] tibble_3.2.1       ggplot2_3.4.4      tidyverse_2.0.0    lintr_3.1.0       
[21] styler_1.10.2      here_1.0.1        

loaded via a namespace (and not attached):
  [1] RColorBrewer_1.1-3     rstudioapi_0.15.0      jsonlite_1.8.7        
  [4] magrittr_2.0.3         spatstat.utils_3.0-4   rmarkdown_2.25        
  [7] vctrs_0.6.4            ROCR_1.0-11            spatstat.explore_3.2-5
 [10] htmltools_0.5.6.1      sctransform_0.4.1      parallelly_1.36.0     
 [13] KernSmooth_2.23-22     htmlwidgets_1.6.2      desc_1.4.2            
 [16] ica_1.0-3              plyr_1.8.9             plotly_4.10.3         
 [19] zoo_1.8-12             igraph_1.5.1           mime_0.12             
 [22] lifecycle_1.0.4        pkgconfig_2.0.3        Matrix_1.6-1.1        
 [25] R6_2.5.1               fastmap_1.1.1          fitdistrplus_1.1-11   
 [28] future_1.33.0          shiny_1.7.5.1          digest_0.6.33         
 [31] colorspace_2.1-0       ps_1.7.5               tensor_1.5            
 [34] rprojroot_2.0.3        RSpectra_0.16-1        irlba_2.3.5.1         
 [37] progressr_0.14.0       fansi_1.0.5            spatstat.sparse_3.0-3 
 [40] timechange_0.2.0       polyclip_1.10-6        abind_1.4-5           
 [43] httr_1.4.7             compiler_4.3.1         remotes_2.4.2.1       
 [46] withr_2.5.2            backports_1.4.1        fastDummies_1.7.3     
 [49] R.utils_2.12.2         MASS_7.3-60            tools_4.3.1           
 [52] lmtest_0.9-40          httpuv_1.6.12          future.apply_1.11.0   
 [55] goftest_1.2-3          R.oo_1.25.0            glue_1.6.2            
 [58] callr_3.7.3            nlme_3.1-163           R.cache_0.16.0        
 [61] promises_1.2.1         grid_4.3.1             Rtsne_0.16            
 [64] cluster_2.1.4          reshape2_1.4.4         generics_0.1.3        
 [67] gtable_0.3.4           spatstat.data_3.0-3    tzdb_0.4.0            
 [70] R.methodsS3_1.8.2      hms_1.1.3              xml2_1.3.5            
 [73] utf8_1.2.4             spatstat.geom_3.2-7    RcppAnnoy_0.0.21      
 [76] ggrepel_0.9.4          RANN_2.6.1             pillar_1.9.0          
 [79] spam_2.10-0            RcppHNSW_0.5.0         later_1.3.1           
 [82] splines_4.3.1          lattice_0.21-8         deldir_1.0-9          
 [85] survival_3.5-7         tidyselect_1.2.0       miniUI_0.1.1.1        
 [88] pbapply_1.7-2          knitr_1.44             gridExtra_2.3         
 [91] scattermore_1.2        xfun_0.40              matrixStats_1.0.0     
 [94] rex_1.2.1              stringi_1.8.1          lazyeval_0.2.2        
 [97] yaml_2.3.7             evaluate_0.22          codetools_0.2-19      
[100] cli_3.6.1              uwot_0.1.16            xtable_1.8-4          
[103] reticulate_1.34.0      munsell_0.5.0          processx_3.8.2        
[106] xmlparsedata_1.0.5     spatstat.random_3.2-1  globals_0.16.2        
[109] png_0.1-8              parallel_4.3.1         ellipsis_0.3.2        
[112] dotCall64_1.1-0        listenv_0.9.0          viridisLite_0.4.2     
[115] cyclocomp_1.1.1        scales_1.2.1           ggridges_0.5.4        
[118] leiden_0.4.3           crayon_1.5.2           rlang_1.1.2      
