---
title: "Psuedobulk DE Analysis"
output: html_document
date: "2024-03-19"
---
The purpose of this script is to run a pseudobulk gene expression analysis using the cell type data from seurat. It is dependent on seurat scripts 01 and 02. It should be run in docker 1.0.8.

Much of this analysis and code is based on the HBC training guide available here: https://hbctraining.github.io/scRNA-seq_online/lessons/pseudobulk_DESeq2_scrnaseq.html

## Load packages

```{r load packages}
suppressPackageStartupMessages({
  library(here)
  library(styler)
  library(lintr)
  library(tidyverse)
  library(patchwork)
  library(ComplexHeatmap)
  library(viridis)
  library(Seurat)
  library(Matrix.utils)
  library(edgeR)
  library(Matrix)
  library(reshape2)
  library(S4Vectors)
  library(SingleCellExperiment)
  library(pheatmap)
  library(apeglm)
  library(png)
  library(DESeq2)
  library(RColorBrewer)
  library(data.table)
})

source(here(
  "src",
  "functions_soelter.R"
))

# set seed
set.seed(42)
```

## Load data

```{r load annotated object}
seurat <-
  readRDS(file = here::here("data", "seurat", "annotated_kidney_samples.rds"))
```

## Prepare data for DESeq2

```{r reformat seurat data to single cell experiment}
# extract raw counts and metadata
counts <- seurat@assays[["RNA"]]@layers[["counts"]]

# get gene names
rownames(counts) <- rownames(seurat)

metadata <- seurat@meta.data

# set up metadata as desired for aggregation and DE analysis
metadata$cluster_id <- factor(metadata$cell_type)

metadata$sample_id <- gsub("_", "", metadata$sample_id)
metadata$sample_id <- factor(metadata$sample_id)

# create single cell experiment object
sce <- SingleCellExperiment(
  assays = list(counts = counts),
  colData = metadata
)
```


```{r get cluster names}
# extract unique names of clusters
cluster_names <- levels(colData(sce)$cluster_id)
cluster_names

# get total number of clusters
length(cluster_names)
```


```{r get sample names}
# extract unique names of samples
sample_names <- levels(colData(sce)$sample_id)
sample_names

# get total number of samples
length(sample_names)
```


```{r get groups}
groups <- colData(sce)[, c("cluster_id", "sample_id")]
head(groups)
```


```{r aggregate expression matrix}
aggr_counts <- aggregate.Matrix(t(counts(sce)),
  groupings = groups, fun = "sum"
)

# explore output matrix
class(aggr_counts)
dim(aggr_counts)
aggr_counts[1:6, 1:6]
```


```{r transpose matrix}
aggr_counts <- t(aggr_counts)
aggr_counts[1:6, 1:6]
```

```{r extract counts for all cell types}
counts_ls <- list()

for (i in seq_along(cluster_names)) {
  ## Extract indexes of columns in the global matrix that match a given cluster
  column_idx <-
    which(tstrsplit(colnames(aggr_counts), "_")[[1]] == cluster_names[i])

  ## Store corresponding sub-matrix as one element of a list
  counts_ls[[i]] <- aggr_counts[, column_idx]
  names(counts_ls)[i] <- cluster_names[i]
}

# Explore the different components of the list
str(counts_ls)
```

```{r get necessary metadata}
# Extract sample-level variables
metadata <- colData(sce) %>%
  as.data.frame() %>%
  dplyr::select(seq_folder, sample_id)

# Exclude duplicated rows
metadata <- metadata[!duplicated(metadata), ]

dim(metadata)
head(metadata)

# restructure for fitting guide better
colnames(metadata) <- c("group_id", "sample_id")

# rename rows
rownames(metadata) <- metadata$sample_id
head(metadata)
```

```{r get cells per cell type for each sample}
t <- table(
  colData(sce)$sample_id,
  colData(sce)$cluster_id
)

t[1:6, 1:6]
```

```{r create metadata list}
## Initiate empty list
metadata_ls <- list()

for (i in seq_along(counts_ls)) {
  ## Initiate a data frame for cluster i
  df <- data.frame(cluster_sample_id = colnames(counts_ls[[i]]))

  ## Use tstrsplit() to separate cluster (cell type) and sample IDs
  df$cluster_id <- tstrsplit(df$cluster_sample_id, "_")[[1]]
  df$sample_id <- tstrsplit(df$cluster_sample_id, "_")[[2]]


  ## Retrieve cell count information for this cluster from cell count table
  idx <- which(colnames(t) == unique(df$cluster_id))
  cell_counts <- t[, idx]

  ## Remove samples with zero cell contributing to the cluster
  cell_counts <- cell_counts[cell_counts > 0]

  ## Match order of cell_counts and sample_ids
  sample_order <- match(df$sample_id, names(cell_counts))
  cell_counts <- cell_counts[sample_order]

  ## Append cell_counts to data frame
  df$cell_count <- cell_counts


  ## Join data frame
  df <- plyr::join(df, metadata,
    by = intersect(names(df), names(metadata))
  )

  ## Update rownames
  rownames(df) <- df$cluster_sample_id

  ## Store complete metadata for cluster i in list
  metadata_ls[[i]] <- df
  names(metadata_ls)[i] <- unique(df$cluster_id)
}

# Explore the different components of the list
str(metadata_ls)
```

## Perform differential gene expression using DESeq2
```{r run deseq2 on all cell types}
# run function on all clusters
map(
  cluster_names,
  get_dds_resultsAvsB,
  A = "wildtype",
  B = "mutant",
  save_path = "results/deseq2/kidney/"
)
```




#### Style

```{r tidy script}
style_file("01_pseudobulk_analysis.Rmd")

lint("01_pseudobulk_analysis.Rmd",
  linters = linters_with_defaults(
    object_length_linter = NULL,
    object_name_linter = NULL,
    object_usage_linter = NULL
  )
)
```

#### Software versions

My software versions will be commented below.

```{r versions}
sessionInfo()
```
R version 4.3.1 (2023-06-16)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 22.04.3 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas-pthread/libblas.so.3 
LAPACK: /usr/lib/x86_64-linux-gnu/openblas-pthread/libopenblasp-r0.3.20.so;  LAPACK version 3.10.0

locale:
 [1] LC_CTYPE=en_US.UTF-8       LC_NUMERIC=C               LC_TIME=en_US.UTF-8       
 [4] LC_COLLATE=en_US.UTF-8     LC_MONETARY=en_US.UTF-8    LC_MESSAGES=en_US.UTF-8   
 [7] LC_PAPER=en_US.UTF-8       LC_NAME=C                  LC_ADDRESS=C              
[10] LC_TELEPHONE=C             LC_MEASUREMENT=en_US.UTF-8 LC_IDENTIFICATION=C       

time zone: Etc/UTC
tzcode source: system (glibc)

attached base packages:
[1] stats4    grid      stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] data.table_1.14.8           RColorBrewer_1.1-3          DESeq2_1.40.2              
 [4] png_0.1-8                   apeglm_1.22.1               pheatmap_1.0.12            
 [7] SingleCellExperiment_1.22.0 SummarizedExperiment_1.30.2 Biobase_2.60.0             
[10] GenomicRanges_1.52.1        GenomeInfoDb_1.36.4         IRanges_2.34.1             
[13] MatrixGenerics_1.12.3       matrixStats_1.0.0           S4Vectors_0.38.2           
[16] BiocGenerics_0.46.0         reshape2_1.4.4              edgeR_3.42.4               
[19] limma_3.56.2                Matrix.utils_0.9.7          Matrix_1.6-1.1             
[22] Seurat_5.0.0                SeuratObject_5.0.0          sp_2.1-1                   
[25] viridis_0.6.4               viridisLite_0.4.2           ComplexHeatmap_2.16.0      
[28] patchwork_1.1.3             lubridate_1.9.3             forcats_1.0.0              
[31] stringr_1.5.1               dplyr_1.1.3                 purrr_1.0.2                
[34] readr_2.1.4                 tidyr_1.3.0                 tibble_3.2.1               
[37] ggplot2_3.4.4               tidyverse_2.0.0             lintr_3.1.0                
[40] styler_1.10.2               here_1.0.1                 

loaded via a namespace (and not attached):
  [1] RcppAnnoy_0.0.21        splines_4.3.1           later_1.3.1            
  [4] bitops_1.0-7            R.oo_1.25.0             polyclip_1.10-6        
  [7] rex_1.2.1               fastDummies_1.7.3       lifecycle_1.0.4        
 [10] doParallel_1.0.17       rprojroot_2.0.3         globals_0.16.2         
 [13] processx_3.8.2          lattice_0.21-8          MASS_7.3-60            
 [16] backports_1.4.1         magrittr_2.0.3          plotly_4.10.3          
 [19] rmarkdown_2.25          yaml_2.3.7              remotes_2.4.2.1        
 [22] httpuv_1.6.12           sctransform_0.4.1       spam_2.10-0            
 [25] spatstat.sparse_3.0-3   reticulate_1.34.0       cowplot_1.1.1          
 [28] pbapply_1.7-2           abind_1.4-5             zlibbioc_1.46.0        
 [31] Rtsne_0.16              R.cache_0.16.0          R.utils_2.12.2         
 [34] RCurl_1.98-1.12         circlize_0.4.15         GenomeInfoDbData_1.2.10
 [37] ggrepel_0.9.4           irlba_2.3.5.1           spatstat.utils_3.0-4   
 [40] listenv_0.9.0           goftest_1.2-3           RSpectra_0.16-1        
 [43] spatstat.random_3.2-1   fitdistrplus_1.1-11     parallelly_1.36.0      
 [46] leiden_0.4.3            codetools_0.2-19        DelayedArray_0.26.7    
 [49] xml2_1.3.5              tidyselect_1.2.0        shape_1.4.6            
 [52] farver_2.1.1            spatstat.explore_3.2-5  jsonlite_1.8.7         
 [55] GetoptLong_1.0.5        ellipsis_0.3.2          progressr_0.14.0       
 [58] ggridges_0.5.4          survival_3.5-7          iterators_1.0.14       
 [61] systemfonts_1.0.4       bbmle_1.0.25.1          foreach_1.5.2          
 [64] tools_4.3.1             ragg_1.2.5              ica_1.0-3              
 [67] Rcpp_1.0.11             glue_1.6.2              gridExtra_2.3          
 [70] xfun_0.40               numDeriv_2016.8-1.1     withr_2.5.2            
 [73] fastmap_1.1.1           fansi_1.0.5             callr_3.7.3            
 [76] digest_0.6.33           timechange_0.2.0        R6_2.5.1               
 [79] mime_0.12               textshaping_0.3.6       colorspace_2.1-0       
 [82] scattermore_1.2         tensor_1.5              spatstat.data_3.0-3    
 [85] R.methodsS3_1.8.2       utf8_1.2.4              generics_0.1.3         
 [88] httr_1.4.7              htmlwidgets_1.6.2       S4Arrays_1.0.6         
 [91] uwot_0.1.16             pkgconfig_2.0.3         gtable_0.3.4           
 [94] lmtest_0.9-40           XVector_0.40.0          htmltools_0.5.6.1      
 [97] dotCall64_1.1-0         clue_0.3-65             scales_1.2.1           
[100] cyclocomp_1.1.1         knitr_1.44              rstudioapi_0.15.0      
[103] tzdb_0.4.0              rjson_0.2.21            coda_0.19-4.1          
[106] nlme_3.1-163            bdsmatrix_1.3-7         zoo_1.8-12             
[109] GlobalOptions_0.1.2     KernSmooth_2.23-22      parallel_4.3.1         
[112] miniUI_0.1.1.1          desc_1.4.2              pillar_1.9.0           
[115] vctrs_0.6.4             RANN_2.6.1              promises_1.2.1         
[118] xtable_1.8-4            cluster_2.1.4           evaluate_0.22          
[121] mvtnorm_1.2-3           cli_3.6.1               locfit_1.5-9.8         
[124] compiler_4.3.1          rlang_1.1.2             crayon_1.5.2           
[127] grr_0.9.5               future.apply_1.11.0     labeling_0.4.3         
[130] emdbook_1.3.13          ps_1.7.5                plyr_1.8.9             
[133] stringi_1.8.1           deldir_1.0-9            BiocParallel_1.34.2    
[136] munsell_0.5.0           lazyeval_0.2.2          spatstat.geom_3.2-7    
[139] RcppHNSW_0.5.0          hms_1.1.3               future_1.33.0          
[142] shiny_1.7.5.1           ROCR_1.0-11             igraph_1.5.1           
[145] xmlparsedata_1.0.5
