---
title: "MARVEL tutorial"
author: "Emma Jones"
date: "2023-10-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## MARVEL Tutorial

```{r load packages}
# Load Lasseigne Lab standard packages
library(here)
library(styler)
library(lintr)

# Load MARVEL package
library(MARVEL)

# Load adjunct packages for selected MARVEL features
# General data processing, plotting
library(ggnewscale)
library(ggrepel)
library(reshape2)
library(plyr)
library(stringr)
library(textclean)

# Gene ontology analysis
library(AnnotationDbi)
library(clusterProfiler)
library(org.Hs.eg.db)
library(org.Mm.eg.db)

# ad hoc gene candidate gene analysis
library(gtools)

# Visualising splice junction location
library(GenomicRanges)
library(IRanges)
library(S4Vectors)
library(wiggleplotr)

# Load adjunct packages for this tutorial
library(Matrix)
library(data.table)
library(ggplot2)
library(gridExtra)
```

```{bash get data}
cd MARVEL_tutorial

wget http://datashare.molbiol.ox.ac.uk/public/project/meadlab/wwen/MARVEL/Tutorial/Droplet/Data.zip

unzip Data.zip

rm Data.zip

```

Now, we want to pull this data into R. I'll start with the normalized gene expression

```{r import normalized gene exp data}
# Matrix
path <- "/data/user/efjones/230926_EJ_Setbp1_AlternativeSplicing/data/MARVEL_tutorial/Data/Gene_SingCellaR/"
file <- "matrix_normalised.mtx"
df.gene.norm <- readMM(paste(path, file, sep = ""))

df.gene.norm[df.gene.norm[, 1] != 0, ][1:5, 1:5]

# cell metadata
path <- "/data/user/efjones/230926_EJ_Setbp1_AlternativeSplicing/data/MARVEL_tutorial/Data/Gene_SingCellaR/"
file <- "phenoData.txt"
df.gene.norm.pheno <- read.table(paste(path, file, sep = ""), sep = "\t", header = TRUE, stringsAsFactors = FALSE)

head(df.gene.norm.pheno)

# gene metadata
path <- "/data/user/efjones/230926_EJ_Setbp1_AlternativeSplicing/data/MARVEL_tutorial/Data/Gene_SingCellaR/"
file <- "featureData.txt"
df.gene.norm.feature <- read.table(paste(path, file, sep = ""), sep = "\t", header = TRUE, stringsAsFactors = FALSE)

head(df.gene.norm.feature)
```

We also want to upload gene counts


```{r import gene counts}
# Matrix
path <- "/data/user/efjones/230926_EJ_Setbp1_AlternativeSplicing/data/MARVEL_tutorial/Data/Gene_STARsolo/"
file <- "matrix_counts.mtx"
df.gene.count <- readMM(paste(path, file, sep = ""))

df.gene.count[df.gene.count[, 1] != 0, ][1:5, 1:5]

# cell metadata
path <- "/data/user/efjones/230926_EJ_Setbp1_AlternativeSplicing/data/MARVEL_tutorial/Data/Gene_STARsolo/"
file <- "phenoData.txt"
df.gene.count.pheno <- read.table(paste(path, file, sep = ""), sep = "\t", header = TRUE, stringsAsFactors = FALSE)

head(df.gene.count.pheno)

# feature metadata
path <- "/data/user/efjones/230926_EJ_Setbp1_AlternativeSplicing/data/MARVEL_tutorial/Data/Gene_STARsolo/"
file <- "featureData.txt"
df.gene.count.feature <- read.table(paste(path, file, sep = ""), sep = "\t", header = TRUE, stringsAsFactors = FALSE)

head(df.gene.count.feature)
```
Looks like the data is all in good shape.

Next, load in splice junction counts.

```{r load in splice junction counts}
# Matrix
path <- "/data/user/efjones/230926_EJ_Setbp1_AlternativeSplicing/data/MARVEL_tutorial/Data/SJ_STARsolo/"
file <- "matrix_counts.mtx"
df.sj.count <- readMM(paste(path, file, sep = ""))

df.sj.count[df.sj.count[, 1] != 0, ][1:5, 1:5]

# cell metadata
path <- "/data/user/efjones/230926_EJ_Setbp1_AlternativeSplicing/data/MARVEL_tutorial/Data/SJ_STARsolo/"
file <- "phenoData.txt"
df.sj.count.pheno <- read.table(paste(path, file, sep = ""), sep = "\t", header = TRUE, stringsAsFactors = FALSE)

head(df.sj.count.pheno)

path <- "/data/user/efjones/230926_EJ_Setbp1_AlternativeSplicing/data/MARVEL_tutorial/Data/SJ_STARsolo/"
file <- "featureData.txt"
df.sj.count.feature <- read.table(paste(path, file, sep = ""), sep = "\t", header = TRUE, stringsAsFactors = FALSE)

head(df.sj.count.feature)
```

Read in remaining files

```{r read in files}
path <- "/data/user/efjones/230926_EJ_Setbp1_AlternativeSplicing/data/MARVEL_tutorial/Data/Gene_SingCellaR/"
file <- "dim_red_coordinates_iPSC_CardioDay10.txt"
df.coord <- read.table(paste(path, file, sep = ""), sep = "\t", header = TRUE, stringsAsFactors = FALSE)

head(df.coord)

path <- "/data/user/efjones/230926_EJ_Setbp1_AlternativeSplicing/data/MARVEL_tutorial/Data/GTF/"
file <- "refdata-cellranger-GRCh38-3.0.0.gtf"
gtf <- as.data.frame(fread(paste(path, file, sep = ""), sep = "\t", header = FALSE, stringsAsFactors = FALSE))
```

Finally, with all this data we can make the MARVEL object

```{r create marvel}
marvel <- CreateMarvelObject.10x(
  gene.norm.matrix = df.gene.norm,
  gene.norm.pheno = df.gene.norm.pheno,
  gene.norm.feature = df.gene.norm.feature,
  gene.count.matrix = df.gene.count,
  gene.count.pheno = df.gene.count.pheno,
  gene.count.feature = df.gene.count.feature,
  sj.count.matrix = df.sj.count,
  sj.count.pheno = df.sj.count.pheno,
  sj.count.feature = df.sj.count.feature,
  pca = df.coord,
  gtf = gtf
)
```

```{r annotate gene metadata}
marvel <- AnnotateGenes.10x(MarvelObject = marvel)

head(marvel$gene.metadata)

table(marvel$gene.metadata$gene_type)
```

```{r annotate junction metadata and validate junctions}
marvel <- AnnotateSJ.10x(MarvelObject = marvel)

head(marvel$sj.metadata)

marvel <- ValidateSJ.10x(MarvelObject = marvel)
```


```{r subset cds genes}
marvel <- FilterGenes.10x(
  MarvelObject = marvel,
  gene.type = "protein_coding"
)
```

## check marvel object

```{r check marvel object}
marvel <- CheckAlignment.10x(MarvelObject = marvel)
```
object is matching and ready for the rest of analysis, so we can save this object

```{r save marvel object}
# Save object
path <- "/data/user/efjones/230926_EJ_Setbp1_AlternativeSplicing/data/MARVEL_tutorial/Data/R Object/"
file <- "MARVEL_EJ.RData"
save(marvel, file = paste(path, file, sep = ""))
```

## Explore expression

```{r gene expression rate}

# Define cell groups
    # Retrieve sample metadata
    sample.metadata <- marvel$sample.metadata
    
    # Group 1 (reference)
    index <- which(sample.metadata$cell.type=="iPSC")
    cell.ids.1 <- sample.metadata[index, "cell.id"]
    length(cell.ids.1)
    # Group 2
    index <- which(sample.metadata$cell.type=="Cardio day 10")
    cell.ids.2 <- sample.metadata[index, "cell.id"]
    length(cell.ids.2)
    
# Explore % of cells expressing genes
marvel <- PlotPctExprCells.Genes.10x(MarvelObject=marvel,
                                     cell.group.g1=cell.ids.1,
                                     cell.group.g2=cell.ids.2,
                                     min.pct.cells=5
                                     )

marvel$pct.cells.expr$Gene$Plot

head(marvel$pct.cells.expr$Gene$Data)
```

```{r differential splicing analysis}

marvel <- CompareValues.SJ.10x(MarvelObject=marvel,
                               cell.group.g1=cell.ids.1,
                               cell.group.g2=cell.ids.2,
                               min.pct.cells.genes=10,
                               min.pct.cells.sj=10,
                               min.gene.norm=1.0,
                               seed=1,
                               n.iterations=100,
                               downsample=TRUE,
                               show.progress=FALSE
                               )

```












```{r tidy script}
style_file("MARVEL_tutorial.Rmd")

lint("MARVEL_tutorial.Rmd",
  linters = linters_with_defaults(
    object_length_linter = NULL,
    object_name_linter = NULL,
    object_usage_linter = NULL
  )
)
```
